\LaTeXe\ guide for the Journal of Functional Programming

### 2019-04-28 papers/jfp/5-rebuilders.tex
Polishing S5

A build system can be split into a scheduler (as defined in \S\ref{sec-scheduler}) and a \emph{rebuilder}. Suppose the scheduler decides that a key should be brought up to date. The next
question is: does any work need to be done, or is the key already up to date?
Or, in a cloud-build system, do we have a cached copy of the value we need?

While \S\ref{sec-background} explicitly listed the schedulers, the rebuilders were introduced more implicitly, primarily by the information they retain to make their decisions. From the examples we have looked at we see four fundamental rebuilds, each with a number of tweaks and variations within them.

@@

The idea of a dirty bit is to have one piece of persistent information per key,
saying whether the key is \emph{dirty} or \emph{clean}. After a build, all bits
are set to clean. When the next build starts, anything that changed between the
two builds is marked dirty. If a key and all its transitive dependencies are
clean, the key does not need to be rebuilt. Taking the example from
Figure~\ref{fig-make}, if \hs{util.c} changes then that would be marked dirty, and \hs{util.o} and \hs{main.exe} would rebuild as they transitively depend on \hs{util.c}.

\Excel models the dirty bit approach most directly, having an actual dirty bit
associated with each cell, marking the cell dirty if the user modifies it.
It also marks dirty all cells that (transitively) depend on the modified cell.
\Excel does not record dynamic dependencies of each cell; instead it computes a
\emph{static over-approximation} -- it is safe for it to make more cells dirty
than necessary, but not vice versa. The over-approximation is as follows: a cell
is marked dirty if its formula statically refers to a dirty cell, or if the
formula calls a function like \cmd{INDIRECT} whose dependencies cannot be
guessed from the formula alone. The over-approximation is clear for
\cmd{INDIRECT}, but it is also present for \cmd{IF}, where both branches are
followed even though dynamically only one is used.

@@

With a dirty bit it is possible to achieve minimality. However, to achieve early
cutoff (\S\ref{sec-background-shake}) it would be important to reset the dirty
bit after a computation that did not change the value and make sure that cells
that depend on it are not rebuilt unnecessarily. For \Excel, this is difficult
because the dependent cells have already been recursively marked dirty. For
\Make, it is impossible to mark a file clean and at the same time not mark the
keys that depend on it dirty. \Make can approximate early cutoff by not
modifying the result file, and not marking it clean, but then it will be rebuilt
in every subsequent build.

The use of a single bit of information is useful to reduce memory consumption, and
in the case of \Make, to integrate with the file system. However, as the examples show,
in constrained environments where a dirty bit is chosen, it's often done as part of
a series of compromises.

@@

An alternative way to determine if a key is dirty is to record the
values/hashes of dependencies used last time, and if something has changed, the
key is dirty and must be rebuilt~--~in essence a \emph{trace} which we can use
to \emph{verify} existing values. Taking the example from Figure \ref{fig-make},
we might record that \hs{util.o} dependend on \hs{util.c} (at hash 1) and \hs{util.h}
(at hash 2) and produced \hs{util.o} (at hash 4). Next time round, if all those files
still have the same hashes, there is nothing to do, and we can skip running the build.
If they are different, we repeat the action, and record the new values.

For traces, there are two essential
operations~--~adding a new value to the trace store, and using the traces to
determine if a key needs rebuilding. Assuming a store of verifying traces
\hs{VT}~\hs{k}~\hs{v}, the operations are:

\begin{minted}[fontsize=\small,xleftmargin=10pt]{haskell}
recordVT :: k -> Hash v -> [(k, Hash v)] -> VT k v -> VT k v
verifyVT :: (Monad m, Eq k, Eq v)
         => k -> Hash v -> (k -> m (Hash v)) -> VT k v -> m Bool
\end{minted}

@@

More interestingly, to \emph{verify} whether a key needs rebuilding we supply
the key, the hash of its current value, a function for obtaining the post-build
value of any key (using a scheduling strategy as per
\S\ref{sec-dependency-orderings}), and the existing \hs{VT} information. The
result will be a \hs{Bool} where \hs{True} indicates that the current value is
already up to date, and \hs{False} indicates that it should be rebuilt.

@@

\subsubsection{Representation Optimisations}

One potential implementation would be to record all arguments passed to
\hs{recordVT} in a list, and verify by simply checking if any list item matches
the information passed by \hs{verifyVT}. Concretely,
in our implementations from \S\ref{sec-implementations}, traces are recorded as lists of:

\begin{minted}[xleftmargin=10pt]{haskell}
data Trace k v = Trace
    {key :: k
    ,depends :: [(k, Hash v)]
    ,result :: Hash v
    }
\end{minted}

\noindent Where \hs{r} is \hs{Hash}~\hs{v} for verifying traces (and \hs{v} for constructive traces, discussed later in \S\ref{sec-constructive-traces}). A real system is highly likely to use a more optimised implementation.

The first optimisation is that any system using \hs{Applicative} dependencies can omit the dependency
keys from the \hs{Trace} since they can be recovered from the \hs{key} field.

The next optimisation is that there is only very minor benefit from storing more than one \hs{Trace} per key. Therefore, veryifying traces can be stored as \hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v}~\hs{(Hash}~\hs{v))}, where the
initial \hs{k} is the \hs{key} field of \hs{Trace} -- making \hs{verifyVT} fast.
Note that storing only one \hs{Trace} per key means that if the dependencies change but the result does not, and then the
dependencies change back to what they were before, there will not be a valid \hs{Trace} available and the system will have to rebuild, where a complete list of all historical traces would not have had to. However, it also means that the number of \hs{Trace} structures is bounded by the number of distinct keys, regardless of how many builds are executed -- a useful property.

>>>>

A build system can be split into a scheduler (as defined
in~\S\ref{sec-scheduler}) and a \emph{rebuilder}. Suppose the scheduler decides
that a key should be brought up to date. The next question is: does any work
need to be done, or is the key already up to date? Or, in a cloud-build system,
do we have a cached copy of the value we need?

While~\S\ref{sec-background} explicitly listed the schedulers, the rebuilders
were introduced more implicitly, primarily by the information they retain to
make their decisions. From the examples we have looked at we see four
fundamental rebuilders, each with a number of tweaks and variations within them.

@@

The idea of a dirty bit is to have one piece of persistent information per key,
saying whether the key is \emph{dirty} or \emph{clean}. After a build, all bits
are set to clean. When the next build starts, anything that changed between the
two builds is marked dirty. If a key and all its transitive dependencies are
clean, the key does not need to be rebuilt. Taking the example from
Fig.~\ref{fig-make}(c), if \cmd{main.c} changes then it would be marked dirty,
and \cmd{main.o} and \cmd{main.exe} would be rebuilt as they transitively depend
on \cmd{main.c}.

\Excel models the dirty bit approach most directly, having an actual dirty bit
associated with each cell, marking the cell dirty if the user modifies it.
It also marks dirty all cells that (transitively) depend on the modified cell.
\Excel does not record dynamic dependencies of each cell; instead it computes a
\emph{static over-approximation} -- it is safe for it to make more cells dirty
than necessary, but not vice versa. The over-approximation is as follows: a cell
is marked dirty (i) if its formula statically refers to a dirty cell, or (ii) if
the formula calls a \emph{volatile} function like \cmd{INDIRECT} whose
dependencies cannot be guessed from the formula alone. The over-approximation is
clear for \cmd{INDIRECT}, but it is also present for \cmd{IF}, where both
branches are followed even though dynamically only one is used.

@@

With a dirty bit it is possible to achieve
minimality~(\S\ref{sec-background-make}). However, to achieve early
cutoff~(\S\ref{sec-background-shake}) it would be important to clear the dirty
bit after a computation that did not change the value and make sure that cells
that depend on it are not rebuilt unnecessarily. For \Excel, this is difficult
because the dependent cells have already been recursively marked dirty. For
\Make, it is impossible to mark a file clean and at the same time not mark the
keys that depend on it dirty. \Make can approximate early cutoff by not
modifying the result file, and not marking it clean, but then it will be rebuilt
in every subsequent build.

A dirty bit rebuilder is useful to reduce memory consumption, and in the case
of \Make, to integrate with the file system. However, as the examples show, in
constrained environments where a dirty bit is chosen, it is often done as part
of a series of compromises.

\q{Andrey}{Can we say something like: ``The authors are not aware of a build
system that uses a dirty bit rebuilder, is minimal, and supports the early
cutoff optimisaion''? Perhaps, we can even make a claim, that such a build
system is impossible with a believable argument/proof sketch?}

@@

An alternative way to determine if a key is dirty is to record the
values/hashes of dependencies used last time, and if something has changed, the
key is dirty and must therefore be rebuilt~--~in essence, keeping a \emph{trace}
which we can use to \emph{verify} existing values. Taking the example from
Fig.~\ref{fig-bazel}(c), we might record that the key \cmd{util.o} (at hash 8)
depended on the keys \cmd{util.c} (at hash 7) and \cmd{util.h} (at hash 2). Next
time round, if the scheduler decides that it is time for \cmd{util.o} to be
rebuilt and all keys still have the same hashes as in the recorded trace, there
is nothing to do, and we can skip rebuilding. If any of the hashes is different,
we rebuild \cmd{util.o}, and record a trace with the new values.

For traces, there are two essential operations~--~adding a new trace to the
trace store, and using the traces to determine if a key needs rebuilding.
Assuming a store of verifying traces \hs{VT}~\hs{k}~\hs{v}, the operations are:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
recordVT :: k -> Hash v -> [(k, Hash v)] -> VT k v -> VT k v
\end{minted}
\vspace{-2mm}
\begin{minted}[xleftmargin=10pt]{haskell}
verifyVT :: (Monad m, Eq k, Eq v)
         => k -> Hash v -> (k -> m (Hash v)) -> VT k v -> m Bool
\end{minted}
\vspace{1mm}

@@

More interestingly, to \emph{verify} whether a key needs rebuilding we use
\hs{verifyVT}, supplying the key, the hash of its current value, a function for
obtaining the post-build value of any key (using a scheduling strategy as
per~\S\ref{sec-scheduler}), and the existing trace store \hs{VT}. The result
will be a \hs{Bool} where \hs{True} indicates that the current value is already
up to date, and \hs{False} indicates that it should be rebuilt.

\q{Andrey}{The "function for obtaining the post-build value of any key" is a bit
mysterious. It would be nice to add a clarification. Or could we simplify the
interface and only provide a function with type
\hs{k -> Hash v -> [(k, Hash v)] -> VT k v -> Bool}? Presumably, \hs{verifyVT}
with the current type signature could be built on top of it?}

@@

\subsubsection{Trace Representation}

One potential implementation would be to record all arguments passed to
\hs{recordVT} in a list, and verify by simply checking if any list item matches
the information passed by \hs{verifyVT}. Concretely, in our implementations
from~\S\ref{sec-implementations}, traces are recorded as lists of:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
data Trace k v a = Trace { key     :: k
                         , depends :: [(k, Hash v)]
                         , result  :: a }
\end{minted}
\vspace{1mm}

\noindent
Where \hs{a} is \hs{Hash}~\hs{v} for verifying traces (and \hs{v} for
constructive traces, discussed later in~\S\ref{sec-constructive-traces}). A real
system is highly likely to use a more optimised implementation.

The first optimisation is that any system using \hs{Applicative} dependencies
can omit the dependency keys from the \hs{Trace} since they can be recovered
from the \hs{key} field~(\S\ref{sec-deps}).

The next optimisation is that there is only very minor benefit from storing more
than one \hs{Trace} per key. Therefore, verifying traces can be stored as
\hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v}~\hs{(Hash}~\hs{v))}, where the
initial \hs{k} is the \hs{key} field of \hs{Trace}, thus making \hs{verifyVT}
fast. Note that storing only one \hs{Trace} per key means that if the
dependencies change but the result does not, and then the dependencies change
back to what they were before, there will be no valid \hs{Trace} available and
the build system will have to rebuild, whereas a complete list of all historical
traces would not have had to. However, it also means that the number of
\hs{Trace} structures is bounded by the number of distinct keys, regardless of
how many builds are executed -- a useful property.
